---
title: "Deaths Norway"
author: "Halvard"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This Rmarkdown contains the analysis of the real data for deaths in Norway and the goal is to fit and compare models with and without adaptive Gaussian Markov random fields in the latent layer of a Bayesian hierarchical model fit through INLA. First, lets introduce the data and perform the necessaru preporssecing. 


# The data - Yearly death rates from Norway
We are specifically interested in periods where we expect shocks. For instance the spanish flu in 1918-1919 ( ref article Tom Borza) and WW2 from 1940-1945 (source?), possibly also from covid. The data is from (source:HMD, and link in sourcelist overleaf), and it is changed and revised periodically, so might be changes. Copied the data from the txt file, and only removed the first line with info about the dataset.

https://www.mortality.org/Country/Country?cntr=NOR

Used the dataset for deaths and 5x1 for ageinterval and yearinterval.
For population I used 5-year, which is yearly and splits age by 5 year intervals. Both are in the column 5x1 and in the section of complete datasets. We then scale the deaths for a age group with the total population in that age group and are left with death rates for each age group. The Spanish flu mainly killed young people, and the same could be said for WW2, as mainly young men went to war and participated in the resistance effort (source maybe?). This is also supported by the data, where we clearly observe the effect of the Spainsh flu and WW2 for the age group $20-24$ while these shocks are less clear for older age groups, say $50-54$. Therefore, I will investigate the death rates for the age $20-24$ from $1900$ to $1970$ with the shocked years $S = \{1918, 1919, 1940, 1941, 1942, 1943, 1944, 1945\}$. 

Maybe a figure of the plots for different ages to illustrate and show that the shocks are visually easy to spot. 

```{r}
DeathsNorway <- read.table("C:/Users/Halvard/Documents/GitHub/Master-project/HMD-data/DeathsNorway.txt", sep = "", header = TRUE, fill = TRUE)

PopulationNorway <- read.table("C:/Users/Halvard/Documents/GitHub/Master-project/HMD-data/PopulationNorway.txt", sep = "", header = TRUE, fill = TRUE)

colnames(DeathsNorway)[5] <- "deaths"
NorwayData <- cbind(DeathsNorway, PopulationNorway[1:4272, "Total"]) #remove the population data for 2024 as there is no deaths data yet

colnames(NorwayData)[6] <- "population"

NorwayData$death_rate <- NorwayData$deaths/NorwayData$population

library(dplyr) #for filter and more
library(ggplot2)
library(ggpubr)
library(INLA)
library(gridExtra)

N <- 71
S_flu <- c(19, 20) # corresponds to 18, 19, the Spanish flu
S_WW2 <- c(41, 42, 43, 44, 45, 46) # corresponds to 40, ..., 45 in years, the WW2
S <- c(S_flu, S_WW2) # all shocked years

col_map <- rep(2, N) #for coloring in the plots below
col_map[S] <- 1

DR20_24 <- filter(NorwayData, Age == "20-24") |>
            filter(Year >= 1900 & Year <= 1970)

plot_20 <- ggplot(DR20_24, aes(Year, death_rate, color = factor(col_map))) + 
          geom_point() +
          labs(x = "Year", y = "Death rate") +
          theme(legend.position = "none")

DR60_64 <- filter(NorwayData, Age == "60-64") |>
            filter(Year >= 1900 & Year <= 1970)

plot_60 <- ggplot(DR60_64, aes(Year, death_rate, color = factor(col_map))) + 
          geom_point() +
          labs(x = "Year", y = "Death rate") +
          theme(legend.position = "none")

plot_20_60 <- ggarrange(plot_20, plot_60, ncol = 2)

death_rate_figure <- annotate_figure(plot_20_60, top = "Death rates for ages 20-24(left) and 60-64(right)")

death_rate_figure

#save the figure
#ggsave("C:/Users/Halvard/Documents/GitHub/Master-project/HMD-plots/death_rate_fig.pdf", plot = death_rate_figure)
```

# Models 

The standard ARW1 with one extra precision for transitions involving shocked years.
```{r}
inla.rgeneric.AdaptiveRW1.model = function(
  cmd = c("graph", "Q", "mu", "initial", "log.norm.const","log.prior", "quit"),
  theta = NULL)
{
  #Input:
  #N is the number of timepoints
  #conflict_years
  #prior_str is either Gamma0.005, Gamma0.00005 or PC
  
  envir = parent.env(environment())
  
  interpret_theta <- function() { return(list(tau1 = exp(theta[1L]), 
                                              tau2 = exp(theta[2L])))}
  
  graph <- function() {return(Q())}
  
  Q <- function() {
    R1 <- matrix(0, nrow = N, ncol = N) #non-conflict
    R2 <- matrix(0, nrow = N, ncol = N) #conflict
    for( i in 1:(N - 1)){
      if(i %in% conflict_years | (i + 1) %in% conflict_years) {
        R2[c(i, i+1), c(i, i+1)] <- R2[c(i, i+1), c(i, i+1)] + c(1, -1, -1, 1)
      }
      else {
        R1[c(i, i+1), c(i, i+1)] <- R1[c(i, i+1), c(i, i+1)] + c(1, -1, -1, 1)
      }
    }
    gv <- exp(1 / N * sum(log(diag(INLA:::inla.ginv(R1 + R2))))) #scaling constant
    R_star_list <- list(R1 = R1*gv, R2 = R2*gv)
    
    p <- interpret_theta()
    Q <- R_star_list$R1 * p$tau1 + R_star_list$R2 * p$tau2
    return(inla.as.sparse(Q)) #sparse representation
  }
  
  mu <- function() {return(numeric(0))}
  
  initial <- function() {return(c(4, 4))}#Default initial for precisions is 4
  
  log.norm.const <- function() {return(numeric(0))}
  
  log.prior <- function() {#default: shape = 1, rate = 0.00005
    p <- interpret_theta()
    if(prior_str == "PC"){
      prior <- inla.pc.dprec(p$tau1, u = 1, alpha = 0.01, log=TRUE) + log(p$tau1) +
          inla.pc.dprec(p$tau2, u = 1, alpha = 0.01, log = TRUE) + log(p$tau2)
      return(prior)
    } else if(prior_str == "Gamma0,005"){
      prior <- dgamma(p$tau1, shape = 1, rate = 0.005, log = TRUE) + log(p$tau1) +
      dgamma(p$tau2, shape = 1, rate = 0.005, log = TRUE) + log(p$tau2)
      return(prior)
    }
    prior <- dgamma(p$tau1, shape = 1, rate = 0.00005, log = TRUE) + log(p$tau1) +
    dgamma(p$tau2, shape = 1, rate = 0.00005, log = TRUE) + log(p$tau2)
    return(prior) 
  }
  
  quit <- function() {return(invisible())}
  
  #to ensure theta is defined
  if (!length(theta)) theta = initial()
  
  vals <- do.call(match.arg(cmd), args = list())
  return(vals)
}
```


The more flexible ARW1 with a separate precision for each shock event. Compared to the previous model we need to pass two lists of conflict points, will assume they do not border each other. Need to make three structure matrices, and also have three parameters in interpret theta and for the log.prior. 
```{r}
inla.rgeneric.AdaptiveRW1.model_3tau = function(
  cmd = c("graph", "Q", "mu", "initial", "log.norm.const","log.prior", "quit"),
  theta = NULL)
{
  #Input:
  #N is the number of timepoints
  #conflict_years1 is the first event
  #conflict_years2 is the second event
  #prior_str is either Gamma0.005, Gamma0.00005 or PC
  
  envir = parent.env(environment())
  
  interpret_theta <- function() { return(list(tau1 = exp(theta[1L]), 
                                              tau2 = exp(theta[2L]),
                                              tau3 = exp(theta[3L])))}
  
  graph <- function() {return(Q())}
  
  Q <- function() {
    R1 <- matrix(0, nrow = N, ncol = N) #non-shocked
    R2 <- matrix(0, nrow = N, ncol = N) #shocked 1
    R3 <- matrix(0, nrow = N, ncol = N) #shocked 2
    for( i in 1:(N - 1)){
      if(i %in% conflict_years1 | (i + 1) %in% conflict_years1) {
        R2[c(i, i+1), c(i, i+1)] <- R2[c(i, i+1), c(i, i+1)] + c(1, -1, -1, 1)
      }
      else if(i %in% conflict_years2 | (i + 1) %in% conflict_years2) {
        R3[c(i, i+1), c(i, i+1)] <- R3[c(i, i+1), c(i, i+1)] + c(1, -1, -1, 1)
      } else{
        R1[c(i, i+1), c(i, i+1)] <- R1[c(i, i+1), c(i, i+1)] + c(1, -1, -1, 1)
      }
    }
    gv <- exp(1 / N * sum(log(diag(INLA:::inla.ginv(R1 + R2 + R3))))) #scaling constant
    R_star_list <- list(R1 = R1*gv, R2 = R2*gv, R3 = R3*gv)
    
    p <- interpret_theta()
    Q <- R_star_list$R1 * p$tau1 + R_star_list$R2 * p$tau2 + R_star_list$R3 * p$tau3
    return(inla.as.sparse(Q)) #sparse representation
  }
  
  mu <- function() {return(numeric(0))}
  
  initial <- function() {return(c(4, 4, 4))}#Default initial for precisions is 4
  
  log.norm.const <- function() {return(numeric(0))}
  
  log.prior <- function() {#default: shape = 1, rate = 0.00005
    p <- interpret_theta()
    if(prior_str == "PC"){
      prior <- inla.pc.dprec(p$tau1, u = 1, alpha = 0.01, log=TRUE) + log(p$tau1) +
          inla.pc.dprec(p$tau2, u = 1, alpha = 0.01, log = TRUE) + log(p$tau2) + 
          inla.pc.dprec(p$tau3, u = 1, alpha = 0.01, log = TRUE) + log(p$tau3)
      return(prior)
    } else if(prior_str == "Gamma0,005"){
      prior <- dgamma(p$tau1, shape = 1, rate = 0.005, log = TRUE) + log(p$tau1) +
               dgamma(p$tau2, shape = 1, rate = 0.005, log = TRUE) + log(p$tau2) +
               dgamma(p$tau3, shape = 1, rate = 0.005, log = TRUE) + log(p$tau3)
      return(prior)
    }
    prior <- dgamma(p$tau1, shape = 1, rate = 0.00005, log = TRUE) + log(p$tau1) +
             dgamma(p$tau2, shape = 1, rate = 0.00005, log = TRUE) + log(p$tau2) +
             dgamma(p$tau3, shape = 1, rate = 0.00005, log = TRUE) + log(p$tau3)
    return(prior) 
  }
  
  quit <- function() {return(invisible())}
  
  #to ensure theta is defined
  if (!length(theta)) theta = initial()
  
  vals <- do.call(match.arg(cmd), args = list())
  return(vals)
}

```


# Fitting the models

## The standard RW1

```{r}
# Defining the model
DR20_24$us <- 1:N #adding columns for the latent models
DR20_24$time <- 1:N
DR20_24$deaths <- as.integer(DR20_24$deaths)
#P <- 1000000000


prior_str <- "PC" #other alternatives are Gamma0,005 and Gamma0,00005 and PC
family <- "poisson"

if(prior_str == "Gamma0,00005"){
  formula_RW1 <- deaths ~ f(time, model = "rw1")#  + f(us, model = "iid")
} else if(prior_str == "Gamma0,005"){
  formula_RW1 <- deaths ~ f(time, model = "rw1", hyper = list(prec = list(prior = "loggamma", param = c(1, 0.005))))#  + f(us, model = "iid")
} else{
  formula_RW1 <- deaths ~ f(time, model = "rw1", hyper = list(prec = list(prior = "pc.prec", param = c(1, 0.01))))#  + f(us, model = "iid")
}

# res_RW1 <- inla(formula_RW1, family = "gaussian", data = DR20_24, 
#                     control.compute = list(cpo = TRUE), 
#                     control.family = list(hyper = list(prec = 
#                                     list(initial = log(P), fixed = TRUE))))

#summary(res_RW1)

#the poisson model
res_RW1 <- inla(formula_RW1, family = family, E = population , data = DR20_24,
           control.compute = list(cpo = TRUE))

summary(res_RW1)
```


## The adaptive RW1 with 2 total precisions

```{r}

ARW1 <- inla.rgeneric.define(inla.rgeneric.AdaptiveRW1.model, 
                      N = N, conflict_years = S, prior_str = prior_str)

formula_ARW1 <- deaths ~ f(time, model = ARW1, extraconstr = 
                  list(A = matrix(1, nrow = 1, ncol = N), e = 0))#  + f(us, model = "iid")

# res_ARW1 <- inla(formula_ARW1, family = "gaussian", data = DR20_24, 
#                     control.compute = list(cpo = TRUE), 
#                     control.family = list(hyper = list(prec = 
#                                     list(initial = log(P), fixed = TRUE))))

#summary(res_ARW1)

#the poisson model
res_ARW1 <- inla(formula_ARW1, family = family, E = population , data = DR20_24,
           control.compute = list(cpo = TRUE))

```


## The adaptive RW1 with 3 total precsions

```{r}
ARW1_3tau <- inla.rgeneric.define(inla.rgeneric.AdaptiveRW1.model_3tau, 
                N = N, conflict_years1 = S_flu, conflict_years2 = S_WW2, prior_str = prior_str)

formula_ARW1_3tau <- deaths ~ f(time, model = ARW1_3tau, extraconstr = 
                    list(A = matrix(1, nrow = 1, ncol = N), e = 0))#  + f(us, model = "iid")

# res_ARW1_3tau <- inla(formula_ARW1_3tau, family = "gaussian", data = DR20_24, 
#                     control.compute = list(cpo = TRUE), 
#                     control.family = list(hyper = list(prec = 
#                                     list(initial = log(P), fixed = TRUE))))

summary(res_ARW1_3tau)
summary(res_RW1)

#the poisson model
res_ARW1_3tau <- inla(formula_ARW1_3tau, family = family, E = population , data = DR20_24,
           control.compute = list(cpo = TRUE))
```

## Show the results with this configuration
lets now try and see how well the models fit the data and calculate the RMSE and LS.

```{r}
#plotting
df <- cbind(DR20_24[, c("Year", "death_rate")], data.frame(eta = res_RW1$summary.fitted.values$mean))

plot_20_RW1 <- ggplot(df, aes(Year)) + 
      geom_point(aes(y = death_rate, color = factor(col_map))) +
      geom_line(aes(y = eta), color = "blue") +  
      labs(x = "Year", y = "Death rate") +
      theme(legend.position = "none") +
      annotate("text", x = 1960, y = max(df$death_rate), label = "RW1",size = 5)

#de to er like
#res_RW1$summary.fitted.values$mean
#res_RW1$summary.linear.predictor$mean

df <- cbind(DR20_24[, c("Year", "death_rate")], data.frame(eta = res_ARW1$summary.fitted.values$mean))

plot_20_ARW1 <- ggplot(df, aes(Year)) + 
      geom_point(aes(y = death_rate, color = factor(col_map))) +
      geom_line(aes(y = eta), color = "blue") +  
      labs(x = "Year", y = "Death rate") +
      theme(legend.position = "none") +
      annotate("text", x = 1960, y = max(df$death_rate), label = "ARW1",size = 5)

df <- cbind(DR20_24[, c("Year", "death_rate")], data.frame(eta = res_ARW1_3tau$summary.fitted.values$mean))

plot_20_ARW1_3tau <- ggplot(df, aes(Year)) + 
      geom_point(aes(y = death_rate, color = factor(col_map))) +
      geom_line(aes(y = eta), color = "blue") +  
      labs(x = "Year", y = "Death rate") +
      theme(legend.position = "none") + 
      annotate("text", x = 1960, y = max(df$death_rate), label = "ARW1_3tau",size = 5)

#make all the three plots together

#RMSE
RMSE1 <- sqrt(mean((DR20_24$death_rate - res_RW1$summary.fitted.values$mean)**2))
RMSE2 <- sqrt(mean((DR20_24$death_rate - res_ARW1$summary.fitted.values$mean)**2))
RMSE3 <- sqrt(mean((DR20_24$death_rate - res_ARW1_3tau$summary.fitted.values$mean)**2))
RMSE <- c(RMSE1, RMSE2, RMSE3)

LS1 <- -mean(log(res_RW1$cpo$cpo))
LS2 <- -mean(log(res_ARW1$cpo$cpo))
LS3 <- -mean(log(res_ARW1_3tau$cpo$cpo))
LS <- c(LS1, LS2, LS3)

data <- data.frame(Method = c("RW1", "ARW1", "ARW1_3tau"), RMSE = RMSE, LS = LS)

ss <- tableGrob(data)

result_plot <- grid.arrange(plot_20_RW1, plot_20_ARW1, plot_20_ARW1_3tau, ss, nrow = 2, top = "With P = 10^9 and PC")

#ggsave("C:/Users/Halvard/Documents/GitHub/Master-project/HMD-plots/res_P10^9_PC.pdf", plot = result_plot, height = 6, width = 8, units = "in")

#INLA::inla.doc("poisson likelihood")

#res_ARW1$cpo$failure
```



### Transforming the parameters from the internal theta scale to standard scale
transforming the marginals for the internal thetas to become precision parameters.
```{r}
exp_func <- function(x){
  return(exp(x))
}

#for RW1
res_RW1$summary.hyperpar

#for ARW1
res_ARW1$summary.hyperpar[3,]
inla.qmarginal(c(0.025, 0.5, 0.975), inla.tmarginal(exp_func, res_ARW1$marginals.hyperpar$`Theta1 for time`))
inla.qmarginal(c(0.025, 0.5, 0.975), inla.tmarginal(exp_func, res_ARW1$marginals.hyperpar$`Theta2 for time`))

#We see that that the iid has the highest precision, then non-shocked RW precision and then the shocked RW precision has the lowest precision

#for ARW1_3tau
res_ARW1_3tau$summary.hyperpar
inla.qmarginal(c(0.025, 0.5, 0.975), inla.tmarginal(exp_func, res_ARW1_3tau$marginals.hyperpar$`Theta1 for time`))



```


# Maybe try without iid term or different priors, maybe also scale the RW1 if it is not the default
















