---
title: "Implementing code from Wakefield for comparison"
author: "Halvard"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
N <- 30
conf_years <- c(9, 10, 11, 12, 13, 14, 15)
prior_str <- "PC" #other alternatives are Gamma0,005 and Gamma0,00005 and PC
if(prior_str == "Gamma0,00005"){
  formula_RW1 <- y ~ f(time, model = "rw1")  + f(us, model = "iid")
} else if(prior_str == "Gamma0,005"){
  formula_RW1 <- y ~ f(time, model = "rw1", hyper = list(prec = list(prior = "loggamma", param = c(1, 0.005))))  + f(us, model = "iid")
} else{
  formula_RW1 <- y ~ f(time, model = "rw1", hyper = list(prec = list(prior = "pc.prec", param = c(1, 0.01))))  + f(us, model = "iid")
}


ARW1_model <- inla.rgeneric.define(inla.rgeneric.AdaptiveRW1.model2, N = N, conflict_years = conf_years, prior_str = prior_str)

formula_ARW1_W <- y ~ f(time, model = ARW1_model,
                extraconstr = list(A = matrix(1, nrow = 1, ncol = N), e = 0)) +
                f(us, model = "iid")

mod_eval_W <- function(df, mean = "", V = "", P, tau = "") {
  df_y <- df[[1]]
  df_eta <- df[[2]]
  n <- dim(df_y)[2] - 2
  
  eval_df <- data.frame(matrix(NA, nrow = n, ncol = 4))
  colnames(eval_df) <- c("RMSE_RW1", "LS_RW1", "RMSE_ARW1", "LS_ARW1")

  for(i in 1:n){#iterate over each simulated realization
    test_data <- df_y[, c(i, n + 1, n + 2)] #gets the i-th realization + time
    colnames(test_data) <- c("y", "time", "us") #makes the colnames match the formula
    
    res_RW1 <- inla(formula_RW1, family = "gaussian", data = test_data, 
                    control.compute = list(cpo = TRUE), 
                    control.family = list(hyper = list(prec = 
                                    list(initial = log(P), fixed = TRUE))))
    LS_RW1 <- -mean(log(res_RW1$cpo$cpo))
    RMSE_RW1 <- RMSE(df_eta[, i], res_RW1$summary.fitted.values$mean)
    
    res_ARW1 <- inla(formula_ARW1_W, family = "gaussian", data = test_data,
                     control.compute = list(cpo = TRUE),
                     control.family = list(hyper = list(prec = 
                                    list(initial = log(P), fixed = TRUE))))
    LS_ARW1 <- -mean(log(res_ARW1$cpo$cpo))
    RMSE_ARW1 <- RMSE(df_eta[, i], res_ARW1$summary.fitted.values$mean)
    
    eval_df[i, ] <- c(RMSE_RW1, LS_RW1, RMSE_ARW1, LS_ARW1)
    
    #Saving the fits
     #save(res_RW1, file=paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/", prior_str, "/Fits_W/RW1/fit_W_", mean, "_", P, "_", tau, "_", i))
     #save(res_ARW1, file=paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/", prior_str, "/Fits_W/ARW1/fit_W_", mean, "_", P, "_", tau, "_", i))
  }
  
  save(eval_df, file=paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/", prior_str, "/Results2/eval_df_", mean, "_", P, "_", tau))
  #save(eval_df, file=paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Harmonius-data/", prior_str, "/eval_df_", mean, "_", P, "_", tau))
  
  #plot_RMSE <- myboxplot2(eval_df[, c("RMSE_RW1", "RMSE_ARW1")], mean, V)
  #plot_LS <- myboxplot2(eval_df[, c("LS_RW1", "LS_ARW1")], mean, V)
  #return(list(RMSE = plot_RMSE, LS = plot_LS))
  return()
}

#example of loading a fit for an ARW1
#load("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/Gamma0,00005/Fits_W/ARW1/fit_W_flat_75_CT_3")
#summary(res_ARW1)
#load("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/Gamma0,00005/Results_W/eval_df_flat_75_CT")
#eval_df
#load("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/Gamma0,00005/Results_W/eval_df_flat_150_CT")
#eval_df
```

```{r}
f_75_CT <- mod_eval_W(data_w_flat_75_CT, "flat", "1/75", 75, "CT")
f_150_CT <- mod_eval_W(data_w_flat_150_CT, "flat", "1/150", 150, "CT")
f_300_CT <- mod_eval_W(data_w_flat_300_CT, "flat", "1/300", 300, "CT")

d_75_CT <- mod_eval_W(data_w_delta_75_CT, "delta", "1/75", 75, "CT")
d_150_CT <- mod_eval_W(data_w_delta_150_CT, "delta", "1/150", 150, "CT")
d_300_CT <- mod_eval_W(data_w_delta_300_CT, "delta", "1/300", 300, "CT")

t_75_CT <- mod_eval_W(data_w_triangle_75_CT, "triangle", "1/75", 75, "CT")
t_150_CT <- mod_eval_W(data_w_triangle_150_CT, "triangle", "1/150", 150, "CT")
t_300_CT <- mod_eval_W(data_w_triangle_300_CT, "triangle", "1/300", 300, "CT")

#Data with non-constant tau
f_75_NCT <- mod_eval_W(data_w_flat_75_NCT, "flat", "1/75", 75, "NCT")
f_150_NCT <- mod_eval_W(data_w_flat_150_NCT, "flat", "1/150", 150, "NCT")
f_300_NCT <- mod_eval_W(data_w_flat_300_NCT, "flat", "1/300", 300, "NCT")

d_75_NCT <- mod_eval_W(data_w_delta_75_NCT, "delta", "1/75", 75, "NCT")
d_150_NCT <- mod_eval_W(data_w_delta_150_NCT, "delta", "1/150", 150, "NCT")
d_300_NCT <- mod_eval_W(data_w_delta_300_NCT, "delta", "1/300", 300, "NCT")

t_75_NCT <- mod_eval_W(data_w_triangle_75_NCT, "triangle", "1/75", 75, "NCT")
t_150_NCT <- mod_eval_W(data_w_triangle_150_NCT, "triangle", "1/150", 150, "NCT")
t_300_NCT <- mod_eval_W(data_w_triangle_300_NCT, "triangle", "1/300", 300, "NCT")
```

Now we splice all the relevant evaluation criteria together with descriptive columnnames. The matrix full_eval_df contains it all.
```{r}
#splicing the eval df's together to one with sensible column names
means <- c("flat", "delta", "triangle")
precs <- c("75", "150", "300")
taus <- c("CT", "NCT")

for(m in 1:length(means)){
  for(p in 1:length(precs)){
    for(t in 1:length(taus)){
      load(paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/", prior_str, "/Results2/eval_df_", means[m], "_", precs[p], "_", taus[t]))
      index_str <- paste0("_", means[m], "_", precs[p], "_", taus[t])
      colnames(eval_df) <- c(paste0("RMSE_RW1", index_str), paste0("LS_RW1", index_str), paste0("RMSE_ARW1", index_str), paste0("LS_ARW1", index_str))
      if(m == 1 & p == 1 & t == 1){
        full_eval_df <- eval_df
      }
      else{
        full_eval_df <- cbind(full_eval_df, eval_df)
      }
    }
  }
}
save(full_eval_df, file=paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/", prior_str,"/Results2/full_eval_df"))
```

Now we need to summarise this information in some plots, and as in the article we will use boxplots for the nine subcases which all have either constant tau or non-constant tau.
```{r}
library(ggpubr) #for plotlist in ggarrange
library(grid)

box_plot_eval <- function(df, mean, P, ymin, ymax, x_text_bool){
  colnames(df) <- c("RW1", "ARW1")
  # Reshape data to long format
  df_long <- df %>%
    pivot_longer(cols = everything(), names_to = "Category", values_to = "Value")
  
  # Create the boxplot
  eval_plot <- ggplot(df_long, aes(x = Category, y = Value, fill = Category)) +
    geom_boxplot() +
    ggtitle(paste0("Mu: ", mean, ", V: 1/", P)) + xlab("") + ylab("") +
    ylim(ymin, ymax) + 
    scale_fill_manual(values = c("RW1" = "skyblue", "ARW1" = "orange")) +
    theme(legend.position ="none") + 
    if(x_text_bool){theme()}
    else{theme(axis.text.x = element_blank())}
  return(eval_plot)
}

make_full_box_plot <- function(df, criteria, tau, prior){
  #Input: df is a full_eval_df, criteria is either "RMSE" or "LS" while tau is either "CT" or "NCT"
  # and prior is the specific prior for the precision
  if(prior == "Gamma0,00005"){prior_name <- "Gamma(1, 0.00005)"}
  else if(prior == "Gamma0,005"){prior_name <- "Gamma(1, 0.005)"}
  else{prior_name <- "PC(1, 0.01)"}
  
  precs <- c("75", "150", "300")
  means <- c("flat", "delta", "triangle")
  plot_list <- list()
  x_text_bool <- FALSE
  for(m in 1:length(means)){
    if(m == length(means)){x_text_bool <- TRUE} #decides if the x_text shows or not
    RW1_str <- paste0(criteria, "_RW1_", means[m], "_") #needs "P_tau"
    ARW1_str <- paste0(criteria, "_ARW1_", means[m], "_") #needs "P_tau"
    df_exempt <- df[, c(paste0(RW1_str, precs[1], "_", tau), paste0(RW1_str, precs[2], "_", tau), paste0(RW1_str, precs[3], "_", tau), paste0(ARW1_str, precs[1], "_", tau), paste0(ARW1_str, precs[2], "_", tau), paste0(ARW1_str, precs[3], "_", tau))] #the relevant columns of df
    
    #Find the min and max value to enforce the same y-axis for the same mean
    ymax <- max(df_exempt) 
    ymin <- min(df_exempt)
    
    for(p in 1:length(precs)){
      plot_list[[(m-1)*3 + p]] <- box_plot_eval(df_exempt[, c(p, p +
      length(precs))],means[m], precs[p], ymin, ymax, x_text_bool)
    }
  }
  plot <- ggarrange(plotlist = plot_list, ncol = 3, nrow = 3)
  #making the title
  title <- paste0(criteria, " for Wakefield data with ")
  if(tau == "CT"){title <- paste0(title, "constant tau and ")}
  else{title <- paste0(title, "non-constant tau and ")}
  title <- paste0(title, prior_name)
  
  plot_w_title <- annotate_figure(plot, top = textGrob(title, gp = (gpar(fontsize = 17))))
  
  ggsave(filename = paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Box-plots/", criteria, "_", tau, "_", prior, ".pdf"), plot = plot_w_title, width = 8, height = 7, units = "in")
  return(plot_w_title)
}
```

Making and saving the plots.
```{r}
load("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/Gamma0,00005/Results2/full_eval_df")
make_full_box_plot(full_eval_df, "RMSE", "CT", "Gamma0,00005")
make_full_box_plot(full_eval_df, "LS", "CT", "Gamma0,00005")
make_full_box_plot(full_eval_df, "RMSE", "NCT", "Gamma0,00005")
make_full_box_plot(full_eval_df, "LS", "NCT", "Gamma0,00005")

load("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/Gamma0,005/Results2/full_eval_df")
make_full_box_plot(full_eval_df, "RMSE", "CT", "Gamma0,005")
make_full_box_plot(full_eval_df, "LS", "CT", "Gamma0,005")
make_full_box_plot(full_eval_df, "RMSE", "NCT", "Gamma0,005")
make_full_box_plot(full_eval_df, "LS", "NCT", "Gamma0,005")

load("C:/Users/Halvard/Documents/GitHub/Master-project/Wakefield-data/PC/Results2/full_eval_df")
make_full_box_plot(full_eval_df, "RMSE", "CT", "PC")
make_full_box_plot(full_eval_df, "LS", "CT", "PC")
make_full_box_plot(full_eval_df, "RMSE", "NCT", "PC")
make_full_box_plot(full_eval_df, "LS", "NCT", "PC")
```


# Harmonius mean evaluation

first, lets get the data, will use the sim_dataframe_sine from Master project.
```{r}
#parameters
N <- 30
n <- 100
tau1 <- 20 #latent precision in the iid noise
tau2 <- 10
o <- 1 # the offset for points 9-15

t <- 1:N
mean_sine <- sin(pi*t/15)

#constant tau and no offset
data_w_harm_75_CT <- sim_dataframe_sine(n, mean_sine, 1/75, tau1, tau1, 0)
data_w_harm_150_CT <- sim_dataframe_sine(n, mean_sine, 1/150, tau1, tau1, 0)
data_w_harm_300_CT <- sim_dataframe_sine(n, mean_sine, 1/300, tau1, tau1, 0)

#constant tau with offset
data_w_sHarm_75_CT <- sim_dataframe_sine(n, mean_sine, 1/75, tau1, tau1, 1)
data_w_sHarm_150_CT <- sim_dataframe_sine(n, mean_sine, 1/150, tau1, tau1, 1)
data_w_sHarm_300_CT <- sim_dataframe_sine(n, mean_sine, 1/300, tau1, tau1, 1)

#non constant tau and no offset
data_w_harm_75_NCT <- sim_dataframe_sine(n, mean_sine, 1/75, tau1, tau2, 0)
data_w_harm_150_NCT <- sim_dataframe_sine(n, mean_sine, 1/150, tau1, tau2, 0)
data_w_harm_300_NCT <- sim_dataframe_sine(n, mean_sine, 1/300, tau1, tau2, 0)

#non constant tau with offset
data_w_sHarm_75_NCT <- sim_dataframe_sine(n, mean_sine, 1/75, tau1, tau2, 1)
data_w_sHarm_150_NCT <- sim_dataframe_sine(n, mean_sine, 1/150, tau1, tau2, 1)
data_w_sHarm_300_NCT <- sim_dataframe_sine(n, mean_sine, 1/300, tau1, tau2, 1)
```

Define the model and the inla formulas used in the modelling.
```{r}
N <- 30
conf_years <- c(9, 10, 11, 12, 13, 14, 15)
prior_str <- "PC" #other alternatives are Gamma0,005 and Gamma0,00005 and PC

if(prior_str == "Gamma0,00005"){
  formula_RW1 <- y ~ f(time, model = "rw1")  + f(us, model = "iid")
} else if(prior_str == "Gamma0,005"){
  formula_RW1 <- y ~ f(time, model = "rw1", hyper = list(prec = list(prior = "loggamma", param = c(1, 0.005))))  + f(us, model = "iid")
} else{
  formula_RW1 <- y ~ f(time, model = "rw1", hyper = list(prec = list(prior = "pc.prec", param = c(1, 0.01))))  + f(us, model = "iid")
}

#R_star_list_W <- Scaled_structure_matrices_for_ARW1(N, conf_years)
ARW1_model_W <- inla.rgeneric.define(inla.rgeneric.AdaptiveRW1.model2, 
                        N = N, conflict_years = conf_years, prior_str = prior_str)
formula_ARW1_W <- y ~ f(time, model = ARW1_model_W,
                extraconstr = list(A = matrix(1, nrow = 1, ncol = N), e = 0)) +
                f(us, model = "iid")
```

Now, lets perform the model evaluation in each case with a given prior.
```{r}
h_75_CT <- mod_eval_W(data_w_harm_75_CT, "harm", "1/75", 75, "CT")
h_150_CT <- mod_eval_W(data_w_harm_150_CT, "harm", "1/150", 150, "CT")
h_300_CT <- mod_eval_W(data_w_harm_300_CT, "harm", "1/300", 300, "CT")

sh_75_CT <- mod_eval_W(data_w_sHarm_75_CT, "sHarm", "1/75", 75, "CT")
sh_150_CT <- mod_eval_W(data_w_sHarm_150_CT, "sHarm", "1/150", 150, "CT")
sh_300_CT <- mod_eval_W(data_w_sHarm_300_CT, "sHarm", "1/300", 300, "CT")

h_75_NCT <- mod_eval_W(data_w_harm_75_NCT, "harm", "1/75", 75, "NCT")
h_150_NCT <- mod_eval_W(data_w_harm_150_NCT, "harm", "1/150", 150, "NCT")
h_300_NCT <- mod_eval_W(data_w_harm_300_NCT, "harm", "1/300", 300, "NCT")

sh_75_NCT <- mod_eval_W(data_w_sHarm_75_NCT, "sHarm", "1/75", 75, "NCT")
sh_150_NCT <- mod_eval_W(data_w_sHarm_150_NCT, "sHarm", "1/150", 150, "NCT")
sh_300_NCT <- mod_eval_W(data_w_sHarm_300_NCT, "sHarm", "1/300", 300, "NCT")

#load(paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Harmonius-data/", "Gamma0,00005", "/eval_df_", "harm", "_", "75", "_", "CT"))

#starta den 16.55
```

Now, lets combine all the evaluation criteria together in the same dataframe.
```{r}
#splicing the eval df's together to one with sensible column names
means <- c("harm", "sHarm")
precs <- c("75", "150", "300")
taus <- c("CT", "NCT")

for(m in 1:length(means)){
  for(p in 1:length(precs)){
    for(t in 1:length(taus)){
      load(paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Harmonius-data/", prior_str, "/eval_df_", means[m], "_", precs[p], "_", taus[t]))
      index_str <- paste0("_", means[m], "_", precs[p], "_", taus[t])
      colnames(eval_df) <- c(paste0("RMSE_RW1", index_str), paste0("LS_RW1", index_str), paste0("RMSE_ARW1", index_str), paste0("LS_ARW1", index_str))
      if(m == 1 & p == 1 & t == 1){
        full_eval_df <- eval_df
      }
      else{
        full_eval_df <- cbind(full_eval_df, eval_df)
      }
    }
  }
}
save(full_eval_df, file=paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Harmonius-data/", prior_str,"/full_eval_df"))
```

Now, lets define the plotting functions to make the boxplot figures with RMSE and LS.

```{r}
library(ggpubr) #for plotlist in ggarrange
# install.packages('latex2exp')
# library(latex2exp)

box_plot_eval_harm <- function(df, mean, P, ymin, ymax, x_text_bool){
  if(mean == "harm"){mean <- "sine"}
  else {mean <- "offset sine"}
  
  colnames(df) <- c("RW1", "ARW1")
  # Reshape data to long format
  df_long <- df %>%
    pivot_longer(cols = everything(), names_to = "Category", values_to = "Value")
  
  # Create the boxplot
  eval_plot <- ggplot(df_long, aes(x = Category, y = Value, fill = Category)) +
    geom_boxplot() +
    ggtitle(paste0("Mu: ", mean, ", V: 1/", P)) + xlab("") + ylab("") +
    ylim(ymin, ymax) + 
    scale_fill_manual(values = c("RW1" = "skyblue", "ARW1" = "orange")) +
    theme(legend.position ="none") + 
    if(x_text_bool){theme()}
    else{theme(axis.text.x = element_blank())}
  return(eval_plot)
}

make_full_box_plot <- function(df, criteria, tau, prior){
  #Input: df is a full_eval_df, criteria is either "RMSE" or "LS" while tau is either "CT" or "NCT"
  # and prior is the specific prior for the precision
  precs <- c("75", "150", "300")
  means <- c("harm", "sHarm")
  if(prior == "Gamma0,00005"){prior_name <- "Gamma(1, 0.00005)"}
  else if(prior == "Gamma0,005"){prior_name <- "Gamma(1, 0.005)"}
  else{prior_name <- "PC(1, 0.01)"}
  
  plot_list <- list()
  x_text_bool <- FALSE
  for(m in 1:length(means)){
    if(m == length(means)){x_text_bool <- TRUE} #decides if the x_text shows or not
    RW1_str <- paste0(criteria, "_RW1_", means[m], "_") #needs "P_tau"
    ARW1_str <- paste0(criteria, "_ARW1_", means[m], "_") #needs "P_tau"
    df_exempt <- df[, c(paste0(RW1_str, precs[1], "_", tau), paste0(RW1_str, precs[2], "_", tau), paste0(RW1_str, precs[3], "_", tau), paste0(ARW1_str, precs[1], "_", tau), paste0(ARW1_str, precs[2], "_", tau), paste0(ARW1_str, precs[3], "_", tau))] #the relevant columns of df
    
    #Find the min and max value to enforce the same y-axis for the same mean
    ymax <- max(df_exempt) 
    ymin <- min(df_exempt)
    
    for(p in 1:length(precs)){
      plot_list[[(m-1)*3 + p]] <- box_plot_eval_harm(df_exempt[, c(p, p +
      length(precs))],means[m], precs[p], ymin, ymax, x_text_bool)
    }
  }
  plot <- ggarrange(plotlist = plot_list, ncol = 3, nrow = 2)
  #making the title
  title <- paste0(criteria, " for Wakefield data with ")
  if(tau == "CT"){title <- paste0(title, "constant tau and ")}
  else{title <- paste0(title, "non-constant tau and ")}
  title <- paste0(title, prior_name)
  
  plot_w_title <- annotate_figure(plot, top = text_grob(title))
  ggsave(filename = paste0("C:/Users/Halvard/Documents/GitHub/Master-project/Harmonius-data/Plots/", criteria, "_", tau, "_", prior, ".pdf"), plot = plot_w_title, width = 8, height = 5, units = "in")
  return(plot_w_title)
}
```

Making and saving the plots.
```{r}
load("C:/Users/Halvard/Documents/GitHub/Master-project/Harmonius-data/Gamma0,00005/full_eval_df")
make_full_box_plot(full_eval_df, "RMSE", "CT", "Gamma0,00005")
make_full_box_plot(full_eval_df, "LS", "CT", "Gamma0,00005")
make_full_box_plot(full_eval_df, "RMSE", "NCT", "Gamma0,00005")
make_full_box_plot(full_eval_df, "LS", "NCT", "Gamma0,00005")

load("C:/Users/Halvard/Documents/GitHub/Master-project/Harmonius-data/Gamma0,005/full_eval_df")
make_full_box_plot(full_eval_df, "RMSE", "CT", "Gamma0,005")
make_full_box_plot(full_eval_df, "LS", "CT", "Gamma0,005")
make_full_box_plot(full_eval_df, "RMSE", "NCT", "Gamma0,005")
make_full_box_plot(full_eval_df, "LS", "NCT", "Gamma0,005")

load("C:/Users/Halvard/Documents/GitHub/Master-project/Harmonius-data/PC/full_eval_df")
make_full_box_plot(full_eval_df, "RMSE", "CT", "PC")
make_full_box_plot(full_eval_df, "LS", "CT", "PC")
make_full_box_plot(full_eval_df, "RMSE", "NCT", "PC")
make_full_box_plot(full_eval_df, "LS", "NCT", "PC")
```









